---
title: "Acquiring Species Data"
format: html
---

```{R}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

library(sf)
library(terra)
library(geodata)
library(mapview)
library(tidyverse)
```

Objectives:

- To get familiar with various sources of online species occurrence data.
- To use relevant R packages to download species occurence data.
- To compare results within the Yukon.

Potential sources to explore:

- GBIF
- eBird
- PlantNet
- iNaturalist
- MoveBank

## GBIF

From their website: [GBIF](https://www.gbif.org/) - the Global Biodiversity Information Facility - is an international network and data infrastructure funded by the world's governments and aimed at providing anyone, anywhere, open access to data about all types of life on Earth.

- View density map of occurrences across [Canada](https://www.gbif.org/country/CA/summary)

- R package: **rgbif**

### Study area

Our example study area is the Kaska Dene Traditional Territory that spans southeast Yukon and northcentral BC. We use the `st_read` function from the **sf** package to read the boundary layer from a geopackage file called kdtt.gpkg. We then use the generic `plot` function to view our study area boundary.

```{r}
#| message: FALSE
#| warning: FALSE
bnd <- st_read('../tutorials/data/kdtt.gpkg', 'bnd', quiet=TRUE)
```

## Moose

We will try to acquire existing moose location data. Sources of species distribution data include the Global Biodiversity Information Facility (GBIF). The `geodata` includes a function to download data from GBIF. The comments in the code below describe the steps to download the data.

```{r}

# Specify genus and species for species of interest (e.g., alces alces for moose)
genus = 'alces'
species = 'alces'

# Download GBIF data and convert to sf point object
g <- sp_occurrence(genus, species, geo=TRUE, end=2500)

# Select a subset of attributes and convert to an sf object
alal <- select(g, lon, lat, year) |>
  st_as_sf(coords = c("lon", "lat"), crs=4326) |>
  st_transform(3578)
cat('Number of observations: ',nrow(alal),'\n')
```

### Map downloaded points

Now we can view our downloaded data.

```{r}
mapview(bnd, alpha.regions=0, lwd=1) + 
  mapview(alal, cex=2, col='red', col.regions='red', legend=NULL)
```

## Caribou

```{r}
#| message: FALSE
#| warning: FALSE
rata <- sp_occurrence('rangifer','tarandus', geo=TRUE, end=2500) |>
  select(lon, lat, year) |>
  st_as_sf(coords = c("lon", "lat"), crs=4326) |>
  st_transform(3578)
cat('Number of observations: ',nrow(rata),'\n')
mapview(bnd, alpha.regions=0, lwd=1) + 
  mapview(rata, cex=2, col='red', col.regions='red', legend=NULL)
```

## Wolverine

```{r}
#| message: FALSE
#| warning: FALSE
gugu <- sp_occurrence('gulo','gulo', geo=TRUE, end=2500) |>
  select(lon, lat, year) |>
  st_as_sf(coords = c("lon", "lat"), crs=4326) |>
  st_transform(3578)
cat('Number of observations: ',nrow(gugu),'\n')
mapview(bnd, alpha.regions=0, lwd=1) + 
  mapview(gugu, cex=2, col='red', col.regions='red', legend=NULL)
```
